---
title: "How do literacy and age of marriage affect family size"
author: "Sakura Hu and Zhanyi Wang"
subtitle: ass1
output:
  pdf_document: default
  html_document: default
---
# Introduction


# Method

# Result

Portugal fertility survey 1979

- [More information](https://wfs.dhsprogram.com/index.cfm?ccode=pt)
- [data source](https://wfs.dhsprogram.com/)
- [data dictionary](https://wfs.dhsprogram.com/pt/ptsr01.dct)

File `portugal.RData` on the course web site.  Code in `Assignment1.Rmd`


```{r dataDownload, include=FALSE}
pUrl = 'http://wfs.dhsprogram.com/pt/ptsr01.dat'
pName = file.path(tempdir(), 'portugal.dat')
if(!file.exists(pName)) {
  download.file(pUrl, pName)
}

datNames = rbind(
		age=c(45,2),
		ageMarried=c(149,2), 
		monthsSinceM = c(157,4),
#		failedPregnancies=c(421,2),
#		failedPregStill=c(423,2),
#		failedPregSpAb=c(425,2),
		pregnancies=c(433,2),
		children=c(435,2),
		sons=c(443,2),
#		firstBirthInterval = c(479,2),
		region = c(641,2),
		literacy = c(649,2)
)
		colnames(datNames ) = c('start','len')
		datNames = cbind(startm1=datNames[,1]-1,datNames, sum=apply(datNames, 1,sum))
		cbind(datNames[-1,1] , datNames[seq(1, nrow(datNames)-1),4])
		datNames[-1,1] = datNames[-1,2] - datNames[seq(1, nrow(datNames)-1),4]
		dWidths = as.vector(t(datNames[,c(1,3)]))
		dNames = paste(rep(rownames(datNames), rep(2, nrow(datNames))),
  	rep(c( "junk",""), nrow(datNames)), sep="") 
		
		dNames = dNames[dWidths > 0]
		dWidths = dWidths[dWidths > 0]
		
		formats = list(
			ageMarried = data.frame(code=1:7,  label=c(0,15,18,20,22,25,30)),
			region = data.frame(code=1:5, 
				label=c('lisbon','porto','20k+', '10-20k', 'lt10k')),
			literacy = data.frame(code=1:2, label=c('yes','no')),
			firstBirthInterval = data.frame(
					code = 1:8,
					label = c(
							'lt0','0-7', '8-11','12-23',
							'24-35','36-47','48-59','60-Inf'
							)
					)
		)

	

		formats$ageMarried$label = 
  	paste(formats$ageMarried$label, 'to',
  	c(formats$ageMarried$label[-1], 'Inf'), sep='')
  	formats$ageMarried = rbind(formats$ageMarried, data.frame(code=88, label='never'))

   
  portugal = read.fwf(
    pName,
    dWidths, col.names=dNames,
    header=FALSE)
  
  portugal = portugal[,grep("junk$", names(portugal), invert=TRUE)]

for(D in intersect(names(portugal), names(formats))){
  		portugal[[D]] = factor(portugal[[D]],
  			levels=formats[[D]]$code, 
				labels=formats[[D]]$label)
}
portugal$ageMarried = relevel(portugal$ageMarried, '22to25')
portugal$region = relevel(portugal$region, 'lt10k')

if(FALSE) save(portugal, file='portugal.RData')
```

```{r thedata}
head(portugal)
table(portugal$region)
```

Figure 1 presents the statistical summary of the response variable, "Children." The number of children per family ranges from 0 to 17, indicating that some families have no children, while others have as many as 17. The distribution is right-skewed, with most families having 2 to 3 children. The mean number of children is 2.26, the median is 2, and the standard deviation is 1.86, reflecting moderate variability. The histogram shows that the data is concentrated around 2 to 3 children, with fewer families having very high numbers of children.

Figure 2 provides the statistical summary of the independent variables, "Age Married" and "Literacy." Among the 5,148 samples, the majority of individuals marry between ages 20 to 25. Specifically, 1,126 individuals married between 20 to 22, and 1,468 individuals married between 22 to 25. In contrast, marriage before age 15 and after 30 is uncommon, with only 52 individuals marrying between 0 to 15 and 217 marrying after 30. The distribution of "Age Married" follows an approximately bell-shaped pattern, slightly left-skewed, with the peak occurring in the 22 to 25 category.

For "Literacy," the majority of individuals are literate. Out of the 5,148 samples, 4,567 individuals reported being literate, while only 581 reported otherwise. The distribution is highly skewed, with literacy being the dominant category.

These statistical summaries provide key insights into the dataset, highlighting the central tendency, spread, and shape of each variable while contextualizing the findings within the sample population.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(grid)
library(knitr)
library(kableExtra)

# Summary statistics for children
children_summary <- portugal %>%
  summarise(
    Mean = round(mean(children, na.rm = TRUE), 2),
    Median = round(median(children, na.rm = TRUE), 2),
    SD = round(sd(children, na.rm = TRUE), 2),
    Min = round(min(children, na.rm = TRUE), 2),
    Max = round(max(children, na.rm = TRUE), 2)
  )

# Create a table with caption
table_with_caption <- arrangeGrob(
  textGrob("Statistical Summary of Children Showing \nMean, Median, Standard Deviation, \nMinimum and Maximum", 
           gp = gpar(fontsize = 10)),
  tableGrob(children_summary),
  heights = c(0.5, 3.0),  # Adjust these values to control spacing
  ncol = 1
)

# Histogram for children with smaller title text
children_hist <- ggplot(portugal, aes(x = children)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.6, color = "black") +
  theme_minimal() +
  labs(title = "Histogram of Number of Children:\nRight-Skewed Distribution with\nMost Families Having 2 to 3 Children", 
       x = "Number of Children", 
       y = "Frequency") +
  theme(plot.title = element_text(size = 10))

# Arrange everything in one figure
grid.arrange(
  table_with_caption,
  children_hist,
  ncol = 2,
  widths = c(1.5, 1),  # Equal widths for both columns
  top = textGrob("Figure 1: Statistical Summary and Histogram of response variable 'Children'",
                 gp = gpar(fontsize = 12, fontface = "bold"))
)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Summary statistics for categorical variables (Age Married and Literacy)
portugal$ageMarried <- factor(portugal$ageMarried, 
                              levels = c("0to15", "15to18", "18to20", "20to22", "22to25", "25to30", "30toInf", "never")) 
ageMarried_summary <- portugal %>%
  count(ageMarried) %>%
  rename(Category = ageMarried, Count = n)

literacy_summary <- portugal %>%
  count(literacy) %>%
  rename(Category = literacy, Count = n)

# Add 'Variable' column and combine the tables
x_var_summary <- bind_rows(
  mutate(ageMarried_summary, Variable = "Age Married"),
  mutate(literacy_summary, Variable = "Literacy")
)

# Reorder the columns to have 'Variable' first
x_var_summary <- x_var_summary[, c("Variable", "Category", "Count")]

# Create a table for X variables summary
x_var_table <- kable(x_var_summary, format = "latex", booktabs = TRUE, 
                      caption = "Statistical Summary of X Variables (Counts)") %>%
  kable_styling(latex_options = c("striped", "hold_position"))

# Bar plot for Age Married
# Define the correct order for ageMarried categories
ageMarried_bar <- ggplot(portugal, aes(x = factor(ageMarried))) +  # Make sure 'ageMarried' is treated as a factor
  geom_bar(fill = "green", alpha = 0.6, color = "black") +
  theme_minimal() +
  labs(title = "Age Married Distribution", x = "Age Married Category", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels to prevent overlap

literacy_bar <- ggplot(portugal, aes(x = literacy)) +
  geom_bar(fill = "orange", alpha = 0.6, color = "black") +
  theme_minimal() +
  labs(title = "Literacy Distribution", x = "Literacy", y = "Count")

# Arrange table & plots in one figure with adjusted widths
grid.arrange(tableGrob(x_var_summary), ageMarried_bar, literacy_bar, ncol = 3, 
             widths = c(1.5, 1.5,1.2),  # Adjust widths for all 3 columns
             top = textGrob("Figure 2: Statistical Summary and Histogram of independnet variables 'Age Married' and 'Literacy'. \nOut of 5148 samples, most samples married between 20 to 25, \n and there are 4567 samples who can read literature",
                            gp = gpar(fontsize = 10, fontface = "bold")))


```
To answer the research question, the data was first modeled using a Poisson regression with an offset for the log-transformed years married. The offset is included to account for the duration of marriage, which may influence the number of children over time. The model is specified as follows:

\[
Y_i \sim \text{Poisson}\left(\exp\left(\text{offset}\left(\log\left(\max(1, \text{monthsSinceM}_i) / 12\right)\right) + \beta_1 \cdot \text{literacy}_i + \beta_2 \cdot \text{ageMarried}_i \right)\right)
\]

Where:
\begin{itemize}
    \item \(\text{children}\) is the count of children,
    \item \(\text{offset}(\log(\max(1, \text{monthsSinceM})/12))\) adjusts for the number of years married,
    \item \(\text{literacy}\) is a binary variable indicating whether the person is literate or not,
    \item \(\text{ageMarried}\) is the categorical variable representing the age at which the person married, with "25to30" as the reference category.
\end{itemize}

Figure 3 shows the summary of this model. According to the results, the variable `literacyno` (indicating illiteracy) has a significant effect on the number of children, with an estimate of 0.159 and a standard error of 0.024. The corresponding z-value is 6.770, and the p-value is less than 0.005, indicating strong evidence against the null hypothesis and  suggesting that being illiterate (compared to being literate) is associated with a higher log count of children. Specifically, being illiterate increases the expected log count of children by 0.159, all else being equal. In contrast, the coefficients for the `ageMarried` categories (relative to the reference group of 25-30 years) do not show significant effects. For example, the coefficient for `ageMarried15to18` is 0.062 with a standard error of 0.037, yielding a z-value of 1.702 and a p-value of 0.089, which is above the standard significance level of 0.05, indicating that this result is not statistically significant. Similarly, other age categories, such as `ageMarried20to22` and `ageMarried30toInf`, show p-values of 0.598 and 0.891, respectively, suggesting no significant relationship between age at marriage and the number of children. Overall, while literacy is a statistically significant predictor, age at marriage does not appear to have a meaningful effect on the number of children in this model.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
portugal$logYearsMarried = log(pmax(1, portugal$monthsSinceM)/12)
portugal$ageMarried = relevel(portugal$ageMarried, "25to30")
portugalFit = glm(children ~ offset(logYearsMarried) + literacy +
ageMarried, data = portugal, family = poisson)
knitr::kable(summary(portugalFit)$coef, digits = 3)
```
Figure 4 presents the means and variances of literacy and age at marriage. In several cases, the variance is at least twice as large as the mean, indicating potential overdispersion. Since the Poisson model assumes equal mean and variance, this violation suggests the need for a more flexible model. To address this, a Negative Binomial model is applied, which allows the variance to exceed the mean by introducing an additional dispersion parameter. The model is specified as follows:
\[
Y_i \sim \text{NegBin}\left(\exp\left(\text{offset}\left(\log\left(\max(1, \text{monthsSinceM}_i) / 12\right)\right) + \beta_1 \cdot \text{literacy}_i + \beta_2 \cdot \text{ageMarried}_i \right), \theta \right)
\]

Where:
\begin{itemize}
  \item \(\text{children}\) represents the number of children,
  \item \(\text{offset}\left(\log\left(\max(1, \text{monthsSinceM}) / 12\right)\right)\) adjusts for the number of years married,
  \item \(\text{literacy}\) is a binary variable indicating whether the person is literate or not,
  \item \(\text{ageMarried}\) is the categorical variable representing the age at which the person married.
\end{itemize}


```{r, echo=FALSE, warning=FALSE, message=FALSE}

m <- aggregate(portugal$children, by=list(portugal$literacy, portugal$ageMarried), mean)

v <- aggregate(portugal$children, by=list(portugal$literacy, portugal$ageMarried), var)

tab<-cbind(m, v[,3])

knitr::kable(tab, digits=2,
col.names=c("literacy",
"agemarried", "Mean", "Variance"))

```

This model shows that the there are The summary of this negative binomial is shown in Figure 5. According to the model results, literacy has a statistically significant effect on the number of children. Specifically, the coefficient for 'literacyno' (indicating illiteracy) is 0.148 with a standard error of 0.027, yielding a z-value of 5.570. The associated p-value is less than 0.05, providing strong evidence against the null hypothesis and suggesting that being illiterate is associated with a higher expected number of children. This suggests that, holding all else constant, being illiterate increases the expected log count of children by 0.148. 

In contrast, the effects of age at marriage do not appear to be statistically significant in this model. The coefficients for different ageMarried categories are estimated relative to the reference group 25–30 years old. with all levels in ageMarried has p value greater than 0.05, which suggest that in this population, the timing of marriage alone may not be a key determinant of fertility outcomes once other factors, such as education, are taken into account. 

The estimated dispersion measure is 0.2645, indicating that the variance exceeds the mean, which violates the Poisson model’s assumption of equidispersion. This justifies the use of a Negative Binomial model, which introduces an additional dispersion parameter to account for the excess variability.


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(MASS)
portugalNB2 <- glm.nb(children ~
offset(logYearsMarried) +
literacy + ageMarried, data = portugal)
knitr::kable(summary(portugalNB2)$coef, digits = 3)
```

```{r}
# Load the necessary library

# Get summary of the model (coefficients and their standard errors)
coef_summary <- summary(portugalNB2)$coefficients

# Get the confidence intervals for the model parameters
conf_intervals <- confint(portugalNB2)

# Display results using knitr::kable
knitr::kable(
  rbind(
    coef_summary[, c(1, 2)],  # Estimates and Standard Errors
    CI_lower = conf_intervals[, 1],  # Lower bound of CI
    CI_upper = conf_intervals[, 2]   # Upper bound of CI
  ),
  caption = "Model Coefficients, Standard Errors, and Confidence Intervals"
)


```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library('glmmTMB')
portugalNB = glmmTMB(
children ~
offset(logYearsMarried) +
literacy + ageMarried, data = portugal, family=nbinom2)

knitr::kable(
  rbind(
    confint(portugalNB)[1:7, c(3, 1, 2)],
    sd = round(1 / sqrt(confint(portugalNB, parm = "sigma"))[3], digits = 3)
  )
)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
exp(confint(portugalNB, 1))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
1/sqrt(confint(portugalNB, parm = "sigma"))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
portugalNBCI = as.data.frame(exp(confint(portugalNB)[-1,]))
portugalNBCI$level = gsub(paste(names(portugalNB$xlevels), collapse='|'),
"", rownames(portugalNBCI))
portugalNBCI$variable=unlist(strsplit(rownames(portugalNBCI),portugalNBCI$level))
portugalNBCI$x = 1:nrow(portugalNBCI)
portugalNBCI$cex = sqrt(1/apply(log(portugalNBCI[,1:2]),1,diff))
forXaxis = tapply(portugalNBCI$x, portugalNBCI$variable, mean)
portugalNBCI
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

# Extract the fixed effects from the model
coef_values <- fixef(portugalNB)$cond

# Apply exp() to the coefficients (excluding the intercept)
exp_coef_values <- exp(coef_values[-1])

# Now you can use 'exp_coef_values' for plotting
matplot(portugalNBCI[, 1:2], type = 'n',
        xaxt = 'n', bty = 'n', log = 'y', ylab = 'RR')
segments(portugalNBCI$x, portugalNBCI[, 1], portugalNBCI$x, portugalNBCI[, 2])
points(portugalNBCI$x, exp_coef_values, pch = 15,
       cex = portugalNBCI$cex, col = '#00000030')
mtext(portugalNBCI$level, 1, at = portugalNBCI$x,
      las = 3, line = -1)
mtext(names(forXaxis), 1, at = forXaxis, line = -2)
abline(h = 1, lty = 3)

```



```{r, echo=FALSE, warning=FALSE, message=FALSE}

1/sqrt(portugalNB2$theta)
```




# Conclusion