Portugal fertility survey 1979

- [More information](https://wfs.dhsprogram.com/index.cfm?ccode=pt)
- [data source](https://wfs.dhsprogram.com/)
- [data dictionary](https://wfs.dhsprogram.com/pt/ptsr01.dct)

File `portugal.RData` on the course web site.  Code in `Assignment1.Rmd`


```{r dataDownload, include=FALSE}
pUrl = 'http://wfs.dhsprogram.com/pt/ptsr01.dat'
pName = file.path(tempdir(), 'portugal.dat')
if(!file.exists(pName)) {
  download.file(pUrl, pName)
}

datNames = rbind(
		age=c(45,2),
		ageMarried=c(149,2), 
		monthsSinceM = c(157,4),
#		failedPregnancies=c(421,2),
#		failedPregStill=c(423,2),
#		failedPregSpAb=c(425,2),
		pregnancies=c(433,2),
		children=c(435,2),
		sons=c(443,2),
#		firstBirthInterval = c(479,2),
		region = c(641,2),
		literacy = c(649,2)
)
		colnames(datNames ) = c('start','len')
		datNames = cbind(startm1=datNames[,1]-1,datNames, sum=apply(datNames, 1,sum))
		cbind(datNames[-1,1] , datNames[seq(1, nrow(datNames)-1),4])
		datNames[-1,1] = datNames[-1,2] - datNames[seq(1, nrow(datNames)-1),4]
		dWidths = as.vector(t(datNames[,c(1,3)]))
		dNames = paste(rep(rownames(datNames), rep(2, nrow(datNames))),
  	rep(c( "junk",""), nrow(datNames)), sep="") 
		
		dNames = dNames[dWidths > 0]
		dWidths = dWidths[dWidths > 0]
		
		formats = list(
			ageMarried = data.frame(code=1:7,  label=c(0,15,18,20,22,25,30)),
			region = data.frame(code=1:5, 
				label=c('lisbon','porto','20k+', '10-20k', 'lt10k')),
			literacy = data.frame(code=1:2, label=c('yes','no')),
			firstBirthInterval = data.frame(
					code = 1:8,
					label = c(
							'lt0','0-7', '8-11','12-23',
							'24-35','36-47','48-59','60-Inf'
							)
					)
		)

	

		formats$ageMarried$label = 
  	paste(formats$ageMarried$label, 'to',
  	c(formats$ageMarried$label[-1], 'Inf'), sep='')
  	formats$ageMarried = rbind(formats$ageMarried, data.frame(code=88, label='never'))

   
  portugal = read.fwf(
    pName,
    dWidths, col.names=dNames,
    header=FALSE)
  
  portugal = portugal[,grep("junk$", names(portugal), invert=TRUE)]

for(D in intersect(names(portugal), names(formats))){
  		portugal[[D]] = factor(portugal[[D]],
  			levels=formats[[D]]$code, 
				labels=formats[[D]]$label)
}
portugal$ageMarried = relevel(portugal$ageMarried, '22to25')
portugal$region = relevel(portugal$region, 'lt10k')

if(FALSE) save(portugal, file='portugal.RData')
```

```{r thedata}
head(portugal)
table(portugal$region)
```

Region is `lt10k` rural areas (less than 10,000 people), towns of size 10-20k, 20k+, and the two largest cities (Lisbon and Porto).


The Question: 

- How do literacy and age of marriage affect family size?
- After we account for known explanatory variables, how much variation (if any) is there in birth rates?

Notes: 

- By European standards Portugal is a poor country, and in 1980 it had the same GDP per capita as Mexico.
- It's well known that families are larger in rural areas (i.e. `region` is a confounder).
- You should proceed as if an expert in this area has told you that you do not need to consider zero-inflation.  Zero-inflated models do not fit well to this data, probably because birth rates are lower than Fiji, many zeros are expected, and the likelihood is flat.

```{r}
table(portugal$ageMarried== '0to15', portugal$literacy)
```


```{r}
# Define groups
literate <- portugal$literacy == "yes"
illiterate <- portugal$literacy == "no"

# Create histogram for literate group
hist(portugal[literate, "children"], 
     breaks = 30, col = rgb(0, 0, 1, 0.5), 
     prob = TRUE, main = "Children vs. Literacy", 
     xlab = "children")

# Add histogram for illiterate group
hist(portugal[illiterate, "children"], 
     breaks = 30, col = rgb(1, 0, 0, 0.5), 
     prob = TRUE, add = TRUE)
     
# Add legend
legend("topright", legend = c("Literate", "Illiterate"), 
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

```
```{r}
portugal$logYearsMarried = log(pmax(1, portugal$monthsSinceM)/12)
portugal$ageMarried = relevel(portugal$ageMarried, "25to30")
portugalFit = glm(children ~ offset(logYearsMarried) + literacy +
ageMarried, data = portugal, family = poisson)
knitr::kable(summary(portugalFit)$coef, digits = 3)
```


```{r}
install.packages("glmmTMB")
```

```{r}
library('glmmTMB')
portugalNB = glmmTMB(
children ~
offset(logYearsMarried) +
literacy + ageMarried, data = portugal, family=nbinom2)

knitr::kable(
  rbind(
    confint(portugalNB)[1:7, c(3, 1, 2)],
    sd = round(1 / sqrt(confint(portugalNB, parm = "sigma"))[3], digits = 3)
  )
)
```

```{r}
exp(confint(portugalNB, 1))
```

```{r}
1/sqrt(confint(portugalNB, parm = "sigma"))
```

```{r}
portugalNBCI = as.data.frame(exp(confint(portugalNB)[-1,]))
portugalNBCI$level = gsub(paste(names(portugalNB$xlevels), collapse='|'),
"", rownames(portugalNBCI))
portugalNBCI$variable=unlist(strsplit(rownames(portugalNBCI),portugalNBCI$level))
portugalNBCI$x = 1:nrow(portugalNBCI)
portugalNBCI$cex = sqrt(1/apply(log(portugalNBCI[,1:2]),1,diff))
forXaxis = tapply(portugalNBCI$x, portugalNBCI$variable, mean)
portugalNBCI[1:3,]
```

```{r}
# Extract the fixed effects from the model
coef_values <- fixef(portugalNB)$cond

# Apply exp() to the coefficients (excluding the intercept)
exp_coef_values <- exp(coef_values[-1])

# Now you can use 'exp_coef_values' for plotting
matplot(portugalNBCI[, 1:2], type = 'n',
        xaxt = 'n', bty = 'n', log = 'y', ylab = 'RR')
segments(portugalNBCI$x, portugalNBCI[, 1], portugalNBCI$x, portugalNBCI[, 2])
points(portugalNBCI$x, exp_coef_values, pch = 15,
       cex = portugalNBCI$cex, col = '#00000030')
mtext(portugalNBCI$level, 1, at = portugalNBCI$x,
      las = 3, line = -1)
mtext(names(forXaxis), 1, at = forXaxis, line = -2)
abline(h = 1, lty = 3)

```

